---
title: "LaScala-Gruenewald algae data"
output: html_document
date: "2023-04-17"
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(vegan)

```

## Limpets grazing on microalgae

These data are from LaScala-Gruenewald et al. 2016, Marine Ecology Progress Series http://doi.org/10.3354/meps11774

The dataset consists of estimates of percent cover of different microalgae groups
growing on experimental plates placed in the intertidal zones. The plates had one
of 4 different herbivore grazing treatments applied, either allowing no grazing,
all natural grazers, or grazing by only members of the limpet species Lottia
scabra, or the species Lottia austrodigitalis. Subsamples from the plates were
preserved for scanning electron microscopy, and the morphotypes of microalgae 
(or bare rubber grip tape on the plates) were visually identified and mean percent
cover estimated based on 27 subsampled locations on each replicate plate. 

Column 1: Site (1-6). This can be treated as a blocking factor. 4 treatment plates were installed together at each Site. 
Column 2: Treatment – the form of grazing treatment present on the plates. Ungrazed plates were surrounded by a fence to exclude all gastropod grazers. Natural plates were open to allow neighboring gastropod grazers to access the plate. Lscabra – plates had 4 Lottia scabra limpets placed within a fence, so grazing should primarily be from that species. Laustro – plates had 4 Lottia austrodigitalis limpets placed within a fence, so grazing should primarily be from that species.
Columns 3-20 represent average percent cover (from 27 subsamples on each plate) of 16 morphotypes of microalgae growing on settlement plates, plus non-biological material or rubber grip tape substrate. Percent cover was calculated based on morphotypes falling under 50 randomly located points


```{r}
micro = read.csv('microalgae_community_SEM_data.csv')
micro$Site = factor(micro$Site)
micro$Treatment = factor(micro$Treatment)
```

```{r}
# Run the nMDS
# micro.nmds = metaMDS(micro[,3:20], distance = 'bray', k = 2, trace = 0)

micro2sqrt = sqrt(sqrt(micro[,3:20])) # Double square root transform of abundances
micro2sqrt = vegdist(micro2sqrt, method = 'bray') # generate distance matrix
# Run the nMDS
micro.nmds = metaMDS(micro2sqrt, distance = 'bray', k = 2, trace = 0,
                     autotransform=FALSE)

micro.nmds  # print info
```

```{r microalgaeStressplot}
# Show the stress plot
stressplot(micro.nmds)
```

```{r plotMicroalgae}
microplot = ordiplot(micro.nmds, type='none', 
		main = paste0('Stress = ',round(micro.nmds$stress,dig=3)),
		xaxt = 'n', yaxt = 'n')
# Plot ellipses first
ordiellipse(micro.nmds,kind='ehull', draw = 'polygon',
		groups=micro$Treatment,label=F, col = 1:4)
# Plot points over ellipses
points(microplot, what = 'sites', pch = as.numeric(micro$Treatment),
       col = as.numeric(micro$Treatment))
# Add legend
legend('topleft',legend=levels(micro$Treatment), col = 1:4,
		pch = 1:4)

```

