---
title: "Final "
author: "Nhi Nguyen"
date: "2025-04-28"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cars}
summary(cars)
```

```{r pressure, echo=FALSE}
plot(pressure)
```

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

install.packages("compositions")       #  CLR transformation
install.packages("ggfortify")  # autoplot pca
install.packages("tidyverse")

library(vegan) # nmd
library(ggplot2) # ggplot
library(dplyr)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(tidyr)
library(compositions)
library(ggfortify)
library(tidyverse)


```

# Introduction

Limpet grazing can strongly influence intertidal microalgae communities. This study explore whether grazing by different limpet species affect the composition and structure of algal communities.

# Methods

```{R}
# Load dataset 
algae_data <- read.csv("microalgae_community_SEM_data.csv")

# Prepare metadata and morphotypes
meta <- algae_data[, c("Site", "Treatment")] # use only the metadata colum (site number and treatment type)
morpho <- algae_data[, 3:ncol(algae_data)] # keep all coulumn from column 3 and so on ( including algae ercent cover data)
algae_all <- cbind(meta, morpho) # combine meta and morpho into 1 dataframe
# metadata (who,what treatment) : non biological info ( site, treatment)
# morpho (what kind) of algae were found, how much of each: include percent cover of each type of microalgae on each plate ( diatom, encrusting , bare space ,..) => what i want to analyze to see how grazing affect algae communities
```

# Bray-Curtis Dissimilarity and nMDS

```{r nmds-analysis}
# calc bray curtis dissimilary base on % cover 
# Bray-Curtis distance
algae_bray <- vegdist(morpho, method = "bray") # see how different each plate is in term of algae composition

# nMDS
# non metric multidimensional scaling using bray distance
algae_nmds <- metaMDS(algae_bray, k = 2, trymax = 100)

# Extract scores from the nmds for each plate ( for plotting)
scores <- as.data.frame(scores(algae_nmds))
# add thje treatment back so I can color point by treatment
scores$Treatment <- meta$Treatment

# make scatter plot for nmds results
# Plot nMDS
ggplot(scores, aes(x = NMDS1, y = NMDS2, color = Treatment)) +
  geom_point(size = 3) +
  stat_ellipse(type = "norm") +
  theme_minimal() +
  labs(title = "nMDS of Microalgae Communities by Treatment")
# each point is a plate , allipse show how similar the groups are

# Stress value= show how well the plot represent the data algae_nmds$stress 
# 0.06 < 0.1 which is very good 
```

# PERMANOVA

```{r}
# permanova
# test to see if comunity composition differs by treatment
# whether algae % cover differ across grazing treatment
adonis_result <- adonis2(morpho ~ Treatment + Site, data = meta, method = "bray")
print(adonis_result) # see the result
# answer the question : Does grazing type affect the algae comm?
# are there location based dofference between site?
```
# Alpha Diversity Analysis

```{R}
# richness is to see how many different taxa 
# shannon is to see both abundance and evenness
#evenness is to see how even the taxa distrbuted
# to see if grazing reduce overall diversity 
# or if ungraze plot support richer, more even communities


# alpha diversity
diversity_data <- algae_all %>% # meta + morpho, creat diversity data with all data + 3 column (richness, shannon, evenness)
  mutate(
    Richness = specnumber(morpho), # count how many algae type are present in each row
    Shannon = diversity(morpho, index = "shannon"),  # how many species , how evenly their abundancde are distribute
    Evenness = Shannon / log(Richness) # cal evennes , standardize shannon by richness. show if species are equally represent or few type dominant
  )

# Plot alpha diversity : richness by treatment
ggplot(diversity_data, aes(x = Treatment, y = Richness ))+
  geom_boxplot(fill = "#FFB3BA") + theme_minimal() +
  labs(title = "Species Richness by Treatment", x = "", y = "Richness (Number of Algal Morphotypes)")
# richness is the number of different algae typpe on each plate
# richness across dif treatment type
# ungraze have highest richness
# graze have lower richness , mean that grazing reduce species variety

# plot alpha diversity : shannon by treatment
ggplot(diversity_data, aes(x = Treatment, y = Shannon)) +
  geom_boxplot(fill = "#D6CDEA") + theme_minimal() +
  labs(title = "Shannon Diversity by Treatment", x = "", y = "Shannon")
# richness + eveness
# ungraze slightly higher diveristy
# graze lower shannon diversity, with some variation
#ungraze higher median = more species , more distribution
# graze show wider spread = diversity is reduce with grazing, low diversity with tougher typoe
# grazing not only reduce diversity but also allow dominance to survive ( skew to tougher , more resitstance tpoe)

# Rarefaction curves
# rarefaction is used to assess sampling effort when dealing with count data
# this data represent percent cover instead of raw species count => rarefaction should not use here. So i focus on richness and evennes using alpha diversity 

```
# CLR- Transformation and PCA

```{R}
# avoid log(0)
morpho_clr <- clr(morpho + 0.001) 
# morpho is data of percent cove , exclude site and treatment
# add 0.001 to avoid log of 0
# clr is centered log-ratio transformation, make the data valid for euclidean distance for PCA

# PCA plot 
pca_clr <- prcomp(morpho_clr, center =TRUE, scale. =TRUE) # run pca on clr transformed data

# prcomp = pca
# center= true is to standardize location
# scale. = true is to standardizes scale

# plot
autoplot(pca_clr, data = meta, colour = 'Treatment') +
  stat_ellipse(aes(group = Treatment), linetype = "dashed") + # add dash ellispses around each treatment group to show clustering
  labs(title = "PCA on CLR-transformed data") +
  theme_minimal() +
  theme(legend.position = "right")

# each point is a plate
# pc1 explain 28.73% , pc2 explain 14.07%
# ungraze plate (purple) is more seperate from other
# graze plate is more overlap, which could mean that they share similar community structure

```


# How algal cover differs across grazing treatments

```{r}
# boxplot

# calc total algae cover by adding all morpho column
algae_all$total_cover <- rowSums(morpho)

# calc total cover for just the filamentous cyanobactereia group
algae_all$Filamentous <- rowSums(algae_all[, c("Filcyano800x150", "Filcyano200x50", "FilcyanoBead10um", "FilcyanoBeadLess10um", "FilcyanoWidthLess7um")], na.rm = TRUE)

# Boxplot 1: Total algae cover across treatments
ggplot(algae_all, aes(x = Treatment, y = total_cover)) +
  geom_boxplot(fill = "lavender") +
  theme_minimal() +
  labs(title = "Total Algal Cover by Treatment", y = "Total % Cover")

# Boxplot 2: Filamentous Group cover across treatment
ggplot(algae_all, aes(x = Treatment, y = Filamentous)) +
  geom_boxplot(fill = "plum1") +
  theme_minimal() +
  labs(title = "Filamentous Cyanobacteria by Treatment", y = "Total % Cover")

# Boxplot 3: a key sensitive morphotype ( filcyano 200x50)
ggplot(algae_all, aes(x = Treatment, y = Filcyano200x50)) +
  geom_boxplot(fill = "lightpink") +
  theme_minimal() +
  labs(title = "Filamentous Cyanobacteria (200x50) by Treatment", y = "Percent Cover")
```

# Stacked Barplot of Relative Composition

```{r}
# Calculate mean % cover of each morpho within each treatment
prop_cover <- algae_data %>%
  group_by(Treatment) %>% # group data by grazing treatment
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>% # calc mean of numeric columns
  pivot_longer(-Treatment, names_to = "Morphotype", values_to = "Cover") # reshape for plotting
# => create new dataframe where each row show the average % cover of specific morpho for each treatment group

# Custom pastel 16 colors because I like these color
my_pastel_colors <- c(
  "#C4E17F", "#C1E1C1", "#ADD8E6", "#FFDAC1", "#FFB347",
  "#E0BBE4", "#D5AAFF", "#FFFFB3", "#FFD1DC", "#B5EAD7",
  "#D6CDEA", "#FF9AA2", "#FFB7B2", "#F9D5E5", "#BFD8B8",
  "#A7D3F2", "#F8B195", "#FAD6A5", "#D0F0C0"
)

#ggplot for stacked boxplot
ggplot(prop_cover, aes(x = Treatment, y = Cover, fill = Morphotype)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = my_pastel_colors) +
  theme_minimal() +
  labs(title = "Relative Morphotype Composition by Treatment", y = "Proportion", x = "")
# show how the community composition shift across treatment. 
# each bar is 1 treatment, colored segment show the relative abundance of each morphotype
```

# Heatmap of Morphotype Cover (extra part to explore treatment-level differences in morphotypes )

```{r}
# calc the averg % cover of each morpho by treatmenbt group 
morpho_means <- algae_data %>%
  group_by(Treatment) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

# prepare the data for heatmap
morpho_only <- morpho_means[,-1] # take out the treatment column
row.names(morpho_only) <- morpho_means$Treatment # make treatment as row
# each row = a grazing treatment
# each column = a morphotyoe
# each cell = mean % cover

# Custom pastel color palette
pastel_palette <- colorRampPalette(c(
  "#FFB3BA", "#B5EAD7","#FFFFB3" , "#D5AAFF", "#BAE1FF"
))(100)
# Generate the heatmap using pastel colors
pheatmap(morpho_only,
         cluster_rows = TRUE, #group similar treatment together
         cluster_cols = TRUE, # group similar morpho together 
         main = "Morphotype Cover by Treatment",
         color = pastel_palette)

```

# Conclusions
```{R}
#Limpet grazing changes what kinds of algae grow on the plates.
#Ungrazed plates had lots of soft, filamentous algae.
#After grazing, those soft types decrease, and tougher, encrusting algae take over.
#The graphs and statistics (like PERMANOVA) all show the same pattern .
```
